---
title: "adoption"
author: "jawwad_kiani"
date: "2025-05-31"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(caret)
library(randomForest)
library(pROC)
library(rpart)
library(e1071)
library(kknn)
library(ggplot2)
library(xgboost)
library(rpart.plot)
library(doParallel)

# Enable parallel processing for faster tuning
cl <- makePSOCKcluster(detectCores() - 1)
registerDoParallel(cl)

# Load adoption data
adoption_data <- read.csv("adoption.csv")

# Clean column names: replace spaces with underscores
names(adoption_data) <- gsub(" ", "_", names(adoption_data))

# Create binary target Adopted (1 if Left_Through_Adoption > 0, else 0)
adoption_data$Adopted <- ifelse(adoption_data$Left.Through.Adoption > 0, 1, 0)

# Select features
features <- c("Male", "Female", "Black", "White", "Hispanic", 
              "Age.Under.1", "Age.1.Through.5", "Age.6.Through.9")

# Filter and prepare data
model_data <- adoption_data %>%
  select(all_of(features), Adopted) %>%
  filter(!is.na(Adopted)) %>%
  na.omit()

# Convert Adopted to factor with levels
model_data$Adopted <- factor(model_data$Adopted, levels = c(0,1), labels = c("No", "Yes"))

# Train-test split
set.seed(123)
train_index <- createDataPartition(model_data$Adopted, p = 0.7, list = FALSE)
train <- model_data[train_index, ]
test <- model_data[-train_index, ]

# Preprocessing (center/scale for some models)
preProc <- preProcess(train[, -ncol(train)], method = c("center", "scale"))
train_scaled <- predict(preProc, train)
test_scaled <- predict(preProc, test)

# Set up repeated CV and tuning control
train_control <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 3,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = TRUE,
  allowParallel = TRUE
)

# Decision Tree tuning grid
dt_grid <- expand.grid(cp = seq(0.001, 0.05, by = 0.005))

# KNN tuning grid (try a range of k)
knn_grid <- expand.grid(kmax = seq(3, 15, 2), distance = 2, kernel = "rectangular")

# XGBoost tuning grid (basic parameters)
xgb_grid <- expand.grid(
  nrounds = c(50, 100),
  max_depth = c(3, 6),
  eta = c(0.1, 0.3),
  gamma = 0,
  colsample_bytree = 0.8,
  min_child_weight = 1,
  subsample = 0.8
)

# Train models

set.seed(123)
models <- list()

# Logistic Regression (scaled data)
models$glm <- train(
  Adopted ~ ., data = train_scaled, method = "glm",
  trControl = train_control, metric = "ROC"
)

# Random Forest (scaled data)
models$rf <- train(
  Adopted ~ ., data = train_scaled, method = "rf",
  trControl = train_control, tuneLength = 5, metric = "ROC"
)

# Decision Tree (unscaled data) - tuned cp
models$rpart <- train(
  Adopted ~ ., data = train, method = "rpart",
  trControl = trainControl(
    method = "cv",
    number = 5,
    classProbs = TRUE,
    summaryFunction = twoClassSummary,
    allowParallel = TRUE
  ),
  tuneGrid = dt_grid,
  metric = "ROC"
)

# KNN (scaled data)
models$knn <- train(
  Adopted ~ ., data = train_scaled, method = "kknn",
  trControl = train_control, tuneGrid = knn_grid, metric = "ROC"
)

# XGBoost (scaled data)
models$xgb <- train(
  Adopted ~ ., data = train_scaled, method = "xgbTree",
  trControl = train_control, tuneGrid = xgb_grid,
  metric = "ROC"
)
```

```{r include= FALSE}
# Evaluate all models
results <- data.frame(
  Model = character(),
  AUC = numeric(),
  Accuracy = numeric(),
  Sensitivity = numeric(),
  Specificity = numeric(),
  stringsAsFactors = FALSE
)

```

```{r include= TRUE}
for (model_name in names(models)) {
  model <- models[[model_name]]
  newdata <- if (model_name %in% c("rpart")) test else test_scaled
  probs <- predict(model, newdata = newdata, type = "prob")[, "Yes"]
  preds <- predict(model, newdata = newdata)
  
  roc_obj <- roc(newdata$Adopted, probs)
  cm <- confusionMatrix(preds, newdata$Adopted, positive = "Yes")
  
  results <- rbind(results, data.frame(
    Model = model_name,
    AUC = auc(roc_obj),
    Accuracy = cm$overall["Accuracy"],
    Sensitivity = cm$byClass["Sensitivity"],
    Specificity = cm$byClass["Specificity"]
  ))
  
  plot(roc_obj, main = paste("ROC Curve -", model_name), col = "blue", lwd = 2)
}
results
```

```{r include= TRUE}
# AUC Comparison Visualization
model_labels <- c(glm = "Logistic Regression", rf = "Random Forest", 
                  rpart = "Decision Tree", knn = "KNN", xgb = "XGBoost")

model_metrics <- results %>%
  mutate(Model = model_labels[Model]) %>%
  arrange(desc(AUC))

ggplot(model_metrics, aes(x = reorder(Model, AUC), y = AUC, fill = Model)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(AUC, 3)), vjust = -0.5) +
  labs(title = "AUC Comparison Across Models", x = "Model", y = "AUC") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r include= TRUE}
# Final Decision Tree Tuning on scaled data
tune_grid <- expand.grid(cp = seq(0.001, 0.05, by = 0.005))

set.seed(123)
dt_tuned <- train(
  Adopted ~ .,
  data = train_scaled,
  method = "rpart",
  trControl = trainControl(
    method = "cv",
    number = 5,
    classProbs = TRUE,
    summaryFunction = twoClassSummary
  ),
  tuneGrid = tune_grid,
  metric = "ROC"
)

print(dt_tuned)
plot(dt_tuned)
```

```{r include= TRUE}
# Prediction using Tuned Decision Tree
dt_probs <- predict(dt_tuned, newdata = test_scaled, type = "prob")[, "Yes"]
dt_preds <- predict(dt_tuned, newdata = test_scaled)
confusionMatrix(dt_preds, test_scaled$Adopted, positive = "Yes")

# ROC and AUC
dt_roc <- roc(test_scaled$Adopted, dt_probs, levels = c("No", "Yes"), direction = "<")
plot(dt_roc, main = "ROC Curve - Tuned Decision Tree", col = "darkred", lwd = 2)
cat("AUC - Tuned Decision Tree:", auc(dt_roc), "\n")
```

```{r include= TRUE}
# Visualize the final tuned decision tree
rpart.plot(dt_tuned$finalModel, type = 2, extra = 104,
           fallen.leaves = TRUE, main = "Final Tuned Decision Tree")
```


### 🧾 Final Model Evaluation Summary (Tuned Decision Tree)

**Confusion Matrix:**
```
            Reference
Prediction   No  Yes
        No   16   9
        Yes   7  18
```

**Performance Metrics:**
- **Accuracy:** 0.68  
- **95% Confidence Interval for Accuracy:** (0.533, 0.8048)  
- **No Information Rate:** 0.54  
- **P-Value [Acc > NIR]:** 0.03133  
- **Kappa:** 0.36  
- **Mcnemar's Test P-Value:** 0.80259  

**Class-wise Metrics:**
- **Sensitivity (Recall for Positive Class):** 0.6667  
- **Specificity (Recall for Negative Class):** 0.6957  
- **Positive Predictive Value (Precision for Yes):** 0.7200  
- **Negative Predictive Value:** 0.6400  
- **Balanced Accuracy:** 0.6812  

**AUC (Area Under ROC Curve):** 0.7303

---

### 🧠 Interpretation:

-  **Tuned Decision Tree** model achieves **moderate performance** with an **AUC of 0.73**, which indicates fair discrimination between "Adopted" and "Not Adopted" cases.
- **Accuracy** is **68%**, which is significantly better than random guessing (NIR = 54%).
- The **Kappa score** of **0.36** indicates a fair level of agreement between predictions and actual outcomes.
- The model shows **balanced sensitivity and specificity**, meaning it handles both classes reasonably well.
- **P-value < 0.05** for accuracy above NIR suggests that the model is statistically better than baseline.
